$Z_3$ topological order in the quantum dimer-pentamer model

### 2017-03-10 qdpm_paper.tex
Incorporating Ty's comment

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.9\linewidth]{example_local_constraints.pdf}
    \caption{Examples of GQDMs with different local constraints. Clockwise from the upper left panel; fixed dimer number: $n_d=1$, $n_d=2$;  fixed dimer parity: odd, even.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link is either unoccupied or occupied by a single dimer. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2+\sigma_\ell^z$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right)
\end{equation}
where $c_P = 2$, ($c_P=1$) for models in the even (odd) sector of the constraint, and the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice. A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires $(Q_v\mod 2)$ to vanish.

@@

However, at the special points in the $\{(t_\nu,v_\nu)\}$ parameter space, known as the Rokhsar-Kivelson (RK) points, when \Eqref{eq:HGQDM} can be written as the sum of projection operators
\begin{equation}
H_{\rm{RK}} = \sum_\nu  {h}_\nu \label{eq:HRK},
\end{equation}
where all $h_\nu$ are projectors with eigenvalues $\{0,1\}$, the exact ground state is known to be
the zero energy state of \Eqref{eq:HRK} that is simultaneously annihilated by all ${\bar{h}}_\nu$. Such a zero energy state, known as the RK state, may be formed from an equal superposition of all dimerizations that are connected by $\{T_\nu\}$:
\begin{equation}
\eket{\Phi_{\RK}^{\Omega}} = \sum_{C \in \Omega} \eket{C}
\end{equation} 
where $\Omega$ is a subset of the Hilbert space connected by $\{T_\nu\}$. Thus $\{ \ket{\Phi_{\RK}^\Omega} \}$ are the exact, degenerate ground states of $H_{\rm{GQDM}}$ with a degeneracy labelled by $\Omega$ at the RK point.

@@

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes $\bplqt$ of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have chosen $t_\bplqt = t$, $v_\bplqt =v, \forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0,
\end{equation}
We can define two conserved topological winding numbers $\Phi_{X}$ \& $\Phi_{Y}$ which are the dimer flux through topologically non-trivial loops that wind around each axis of the lattice when the system has periodic boundary conditions (see \figref{fig:orientedSL}). On a lattice of dimensions $L_x\times L_y$, the eigenvalues of $\Phi_X$ and $\Phi_Y$ are integers in the range
\begin{equation}
-L_{x}/2 \leq  \Phi_{X} \leq L_{x}/2,
\end{equation}
and similarly for $\Phi_Y$, thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

At the Rokhsar-Kivelson point $t=v$, we may write $\HQDM$ in the form of \Eqref{eq:HRK} by defining projection operators  $ {h}_\nu \equiv  {P}^\dagger_{\nu} {P}_{\nu}$, with
\begin{equation}
 {P}^\dagger_\nu \equiv \biggl( \eket{ c_\nu} - \eket{\bar{c}_\nu} \biggr).
\end{equation} 
Since the plaquette flip term in \Eqref{eq:HQDM} is ergodic in each of topological sector $(\Phi_x,\Phi_y)$ on the square lattice~\cite{????}, each topological sector corresponds to a unique RK ground state where $\Omega$ is labeled by $(\Phi_x,\Phi_y)$. Thus the  RK points of the QDM and QFPLM have an \emph{extensively} degenerate ground states.

@@

The increased dynamics of $\HIGT$ vs. $\HQDM$ no longer conserve the dimer flux $\Phi_\Gamma$ through a closed curve; only $\Phi_\Gamma \mod 2$ is preserved by $\HIGT$. Consequently, to define the string and loop operators that commute with $\HIGT$ (for closed strings) we fix $\alpha = \pi$ in \Eqref{eq:Sgamma}:
\begin{equation}
S_\gamma^{Z_2} \equiv S_\gamma \left( \pi \right), \quad W_{X,Y}^{Z_2} \equiv W_{X,Y} \left( \pi \right).
\end{equation}
Thus the $W_{X,Y}^{Z_2}$ have two eigenvalues $\pm 1$, and the topological sectors on the torus can be labeled as~$\Omega= (\pm 1,\pm1)$; consequently the extensive degeneracy of the fixed number constrained models has been reduced to a finite ground state degeneracy in the fixed parity models.

@@

We have demonstrated that ground state at the RK point of the QDPM has an exact $Z_3$ gauge symmetry and a corresponding $9$-fold topological degeneracy on the torus. We emphasize that despite the exact $Z_3$ local gauge symmetry of the QDPM, there is not an exact mapping to a $Z_3$ gauge theory on the square lattice, as the QDPM has Ising degrees of freedom on the links, rather than the 3 dimensional local Hilbert space of a $Z_3$ gauge theory (see \ref{sec:Z3GT} for details of $Z_3$ gauge theory).  It remains to address whether this RK state is in the deconfined $Z_3$ topologically ordered phase of a $Z_3$ gauge theory, or a confined symmetry broken or critical state, which we discuss below.

@@

Computations were performed on the Vermont Advanced Computing Core supported by NASA (NNX-08AO96G).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.9\linewidth]{example_local_constraints.pdf}
    \caption{Examples of GQDMs with different local constraints. Clockwise from the upper left panel; fixed dimer number: $n_d=1$, $n_d=2$;  fixed dimer parity: odd, even.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link is either unoccupied or occupied by a single dimer. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2+\sigma_\ell^z$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space is spanned by an orthonormal basis of dimerizations of the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right)
\end{equation}
where $c_P = 2$, ($c_P=1$) for models in the even (odd) sector of the constraint, and the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice. A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires $(Q_v\mod 2)$ to vanish.

@@

However, at the special points in the $\{(t_\nu,v_\nu)\}$ parameter space, known as the Rokhsar-Kivelson (RK) points, when \Eqref{eq:HGQDM} can be written as the sum of projection operators
\begin{equation}
H_{\rm{RK}} = \sum_\nu  {h}_\nu \label{eq:HRK},
\end{equation}
where all $h_\nu$ are projectors with eigenvalues $\{0,1\}$, the exact ground state is known to be
the zero energy state of \Eqref{eq:HRK} that is simultaneously annihilated by all ${h}_\nu$. Such a zero energy state, known as the RK state, may be formed from an equal superposition of all dimerizations that are connected by $\{T_\nu\}$:
\begin{equation}
\eket{\Phi_{\RK}^{\Omega}} = \sum_{C \in \Omega} \eket{C}
\end{equation} 
where $\Omega$ is a subset of the Hilbert space connected by $\{T_\nu\}$. Thus $\{ \ket{\Phi_{\RK}^\Omega} \}$ are the exact, degenerate ground states of $H_{\rm{GQDM}}$ with a degeneracy labelled by $\Omega$ at the RK point.

@@

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes $\bplqt$ of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have chosen $t_\bplqt = t$, $v_\bplqt =v, \forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0.
\end{equation}
We can define two conserved topological winding numbers $\Phi_{X}$ \& $\Phi_{Y}$ which are the dimer flux through topologically non-trivial loops that wind around each axis of the lattice when the system has periodic boundary conditions (see \figref{fig:orientedSL}). On a lattice of dimensions $L_x\times L_y$, the eigenvalues of $\Phi_X$ and $\Phi_Y$ are integers in the range
\begin{equation}
-L_{x}/2 \leq  \Phi_{X} \leq L_{x}/2,
\end{equation}
and similarly for $\Phi_Y$, thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

At the Rokhsar-Kivelson point $t=v$, we may write $\HQDM$ in the form of \Eqref{eq:HRK} by defining projection operators  $ {h}_\nu $:
\begin{equation}
 {h}_\nu \equiv \biggl( \eket{ c_\nu} - \eket{\bar{c}_\nu} \biggr)\biggl( \ebra{ c_\nu} - \ebra{\bar{c}_\nu} \biggr).
\end{equation} 
Since the plaquette flip term in \Eqref{eq:HQDM} is ergodic in each of topological sector $(\Phi_x,\Phi_y)$ on the square lattice~\cite{????}, each topological sector corresponds to a unique RK ground state where $\Omega$ is labeled by $(\Phi_x,\Phi_y)$. Thus the  RK points of the QDM and QFPLM have an \emph{extensively} degenerate ground states.

@@

The additional dynamics of $\HIGT$ vs. $\HQDM$ no longer conserve the dimer flux $\Phi_\Gamma$ through a closed curve; only $\Phi_\Gamma \mod 2$ is preserved by $\HIGT$. Consequently, to define the string and loop operators that commute with $\HIGT$ (for closed strings) we fix $\alpha = \pi$ in \Eqref{eq:Sgamma}:
\begin{equation}
S_\gamma^{Z_2} \equiv S_\gamma \left( \pi \right), \quad W_{X,Y}^{Z_2} \equiv W_{X,Y} \left( \pi \right).
\end{equation}
Thus the $W_{X,Y}^{Z_2}$ have two eigenvalues $\pm 1$, and the topological sectors on the torus can be labeled as~$\Omega= (\pm 1,\pm1)$; consequently the extensive degeneracy of the fixed number constrained models has been reduced to a finite ground state degeneracy in the fixed parity models.

@@

We have demonstrated that ground state at the RK point of the QDPM has an exact $Z_3$ gauge symmetry and a corresponding $9$-fold topological degeneracy on the torus. We emphasize that despite the exact $Z_3$ local gauge symmetry of the QDPM, there is not an exact mapping to a $Z_3$ gauge theory on the square lattice, as the QDPM has Ising degrees of freedom on the links, rather than the 3 dimensional local Hilbert space of a $Z_3$ gauge theory (see \appref{sec:Z3GT} for details of $Z_3$ gauge theory).  It remains to address whether this RK state is in the deconfined $Z_3$ topologically ordered phase of a $Z_3$ gauge theory, or a confined symmetry broken or critical state, which we discuss below.

@@

We thank T. J. Volkoff for discussions and a careful reading of the manuscript. Computations were performed on the Vermont Advanced Computing Core supported by NASA (NNX-08AO96G).
